Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-09-01 Scripts/Manuscript.Rmd
Methods completed for first draft

The delineation of Economic Exclusive Zones (EEZs) by the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, established political boundaries across the distribution of many fish species, creating shared species between nations. Despite not implicit, UNCLOS incentivizes nations to cooperate in the management of transboundary fisheries and most management plans emplaced are based on historical distribution of catches. Climate change represents a challenge for the joint management of fish stocks. For instance, climate-driven changes in species distributions have led to conflict between nations due to changes in the proportion of captures, quota allocation, and fishery newcomers. Recent events suggests that even changes in species that contribute a relatively small proportion of the overall catch can lead to important international conflict when no legal framework is available. Thus, it is fundamental to prepare for cliamte change-driven impacts such as identifying the stocks that are going to change historical proportions within neighbouring EEZs, the degree by which they will cahnge, and the time when such change is going to happen. Only by adapting management plans and polocy that antecipates to these effects will international fisheries treaties be climate-ready. Here, we build on previous work to identify the transboundary stocks that will see changes in their stock-share-proportion (e.g. the proportion of the stock that each EEZ has) around the world. Moreover, we determine the time of emergence of each fish stock, defined as the year in which the stock-share-proportion will change from its historical range. Our results suggest that by the mid of the 21st century EEZs will see changes in stock-share ratio between 1 and 10% with some exceeding 25%. Most of these changes will hapen before 2050 urging policy to be re-formulated to account for the possible implications of these changes. By the end of the 21st century virtually all EEZs and transboundary sepeices will have experienced changes in stock-share-proportion relative to historical distributions. The results of this research has the potential to improve climate-ready fisheries if policy is modify to respond to the consequences presented here.

@@

<!-- The world is moving towards sustainable development and ocean-based activities plays a mayor role in achieveing international policies. -->

<!-- One of the main goals of the United Nations Sustainable Development Goal No.14 (SDG 14-Life below water) is to achieve sustainable development of fisheries [@SDG14]. -->
<!-- (https://sustainabledevelopment.un.org/sdg14). -->

<!-- Climate change is expected to reduce the production from global fisheries in about XX% to xx% [@Cheung; @Barange] compromisisng our capacity to reach SDG 14. -->

<!-- Recent reseach show that the impacts of climate change on the productivity of marine wild fisheries can be offset with the proper reform of current fisheries management [@Gaines; @CheungClimateReadyFish]. While bringing fisheries that are poorly managed to a sustainable level will likeley offset the negative impacts of climate change to fisheries production [@Gaines], it will not necessarilly solve the spatial consequences of species moving from one mangement area to another.  -->

<!-- Shifts in species distribution does not necessarilly mean a reduction in production (e.g. reduction in stock biomass) but rather a spatial re-distribution of the stock's biomass [@RefLiam?].  -->

<!-- Documented cases of stock shifts have resulted in international conflict and usustainable fishing rates by parties involved in the conflict. -->

<!-- This is of specific concern to a set of 633 fished species that cross international borders, also known as transbounday species [@FishForVisa].  -->

<!-- With the implementation of arbritary spatial management boundaries, at least 633 species became transboudary representing an average catch (2005 - 2010) of 47.6 million tons and generating USD 65,483 million in fishing revenue [@FishorVisa].  -->

<!-- Transboudary fish stocks refere to those species that cross international jurisdictions known as Economic Exclusive Zones (EEZs) of neighbouring nations [@FAO]. These stocks are responsable for an average catch (2005 - 2010) of 47.6 million tons generating USD 65,483 million in fishing revenue [@FishorVisa]. -->

<!-- Historically, environmentally-driven shits in shared stocks have cause problems and threathen stock's sustainabillity. **In early 1980's Canada and the US blablabla.** Many fish stocks managed by the European Union ICES regultaroy areas have been shifting since the 1970's [@EuropeShift]. Such shifts have already result in confrontations between nations as evidenced by the expansion of the Atlantic mackerel (*Scomber scombrus*) distribution reaching the EEZ of Iceland [@Spijkers:2017ij].  More recently, Chile and Peru have declare a cooperation strategy to mange the transboundary stock of Peruvean Anchovera [@Peru]. Furthermore, as we continue to enter the $21^{st}$ century, commercial marine species are expected to continue shifting (mostly poleward) potentilly compromising shared stocks management. By the end of the century, all EEZs of the world are proyected to have at least one new transboundary species with some of them having over 10 new stocks [@Pinsky]. Consequently, as species occupy new waters they leave historic areas and so, xx% of species are expected to exit EEZs by the end of che century [@Oremus]. This entering and exicting national jurisdicions will complicate fisheries management potentially creating confllict between neighbouring nations [@Pinksy @Oremus:2020id]. -->

<!--  So, preparing policy to species on the move is key to achieve sustainable development goals and proper governance of the world fisheries [@Pinksy]. -->

<!-- Undrstanding the degree by which shared stocks are going to change and the time of emergence is key for creating climate-ready fisheries policy. -->

<!-- <!-- The concepts of transboundary stocks and EEZs were introduced in the early 1980's by the United Nations Law of the Sea (UCLOS) [@UnitedNations:1986tl]. Under UNCLOS each costal nation has the right (and responsabillity) to manage wild fish stocks up to 200 nautical mailes of the coast. Exceptions exist when territorial waters are in dispute (e.g. South China sea[@Ref]) or when menagement agreements exist between nations [@Ref]. In fact, UNCLOS incentives actions to cooperate on the management of shared stocks [Article 63, @UnitedNations:1986tl]. Climate change is shifting the distribution of marine species to the extreem that some spatial policies implemented to manage fisheries are becoming obsolete [@Pololanza; @Jessica].  --> -->

@@

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hearafter called DBEM). The DBEM algorythm includes species physiology, habitat suitabillity, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario accroding to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variabillity, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensambles the Earth system initial condition is infinitesimally different resulting in an unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensamble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variabillity (e.g. backround noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distributio by averaging the secular trend over all ensemble members [@Rodggers:2015jq].

## Data analysis

The Food and Agricultural Organization (FAO) of the UN define transboundary stocks as those that freely swim across the EEZ of neighbouring nations [@REFFAO]. Accordingly, we included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs [@Palacios-Abrantes:xxxx]. We consider each species as a single stock unit within an EEZ due to lack of more detailed information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the Sea Around Us shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp].

### Changes in Stocks Share Ratio

In game theory, a threat point is known as the minimum payoff that a player is willing to receive in order to cooperate with other players. Thus, if a player's payoff is lower than the theat point, such player will liekly break off any collaboration agreement and proceed to act solo [@Sumaila:E&S]. Here, we adopted the concept of threat point as the minimum proportion of the catch that each country will acept if they were to cooperate in the management of a transboundary stock [@Palacios-Abrantes:E&S] hereafter referred as stock share ratio (SSR).  For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighbouring EEZs, and then calculated the proportion that each EEZ had [@Palacios-Abrantes:xxxx]. We then average the results in three time periods in order to reduce climate variabillity within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighbouring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^st$ century as the average of 2021 to 2040 ($t_e$), and mid $21^st$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. We then estimate the percentage change in sthe SSR of each future time period relative to the historic time period. To address the natural spatal variabillity of the stock we only considered a SSR chgange when the future SSR was higher or lower than the average historic SSR $\pm$ 1 s.d. The analysis was repplicated for each of the ten ensamble members capturing variation by averaging results among ensamble members (mean $\pm$ standar deviation).

### Time of Emergence Index

Knowing the time (e.g. year) in whitch SSR is going to diverge from a historical variablillty can be key for policy to prepare to such change. In here we call this event time of emergence and defined it as the moment in time when the future variation in SSR becomes larger than the historical trend variabillity [@Roggers:2015jq]. We employed the ten GFDL ensable members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in realtion to the centroid of the neighbouring EEZs it ocuppys. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensamble member, neighbouring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences:

$$D_{ens} = acos(sin(lat_{eez})*sin(lat_{spp}) + cos(lat_{eez})*cos(lat_{spp})*cos(lon_{eez}-lon_{spp}))$$
where $lat_{eez}$ and $lat_{spp}$ are the latitudes of the EEZ and specie centroid, respectiveley and $lon_{eez}$ and $lon_{spp}$ are the respectevly longitudes.

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectiveley for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*h*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, to determine the time of emergence we estimated a signal to noise ratio (SNR) and set an arbitrary threshold as follows [@Rodggers:2015jq]:

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standar deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbritary decission, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated following the previos equation.

<!-- In cases where the species is distributed in more than 2 neighbouring EEZs (e.g. *xxx*), the index is as follows: -->

@@

In other words, for two neighbouring EEZs sharing a stock, if the centroid shifts towards EEZ A ($D_A$ decreases) but away from EEZ A B ($D_B$ increases), EEZ A may catch a bigger share over B and the risk of dispute increases (see *TI* equation). Consequently, if the centroid shifts are within historical variability (*SNR* < 1), risk of dispute between EEZ A and B is lower, finally, if the centroid shifts are beyond historical variability (*SNR* > 1), the risk of dispute is higher. Thus, risk is positively related to the frequency of centroid shifts beyond historical variability.

We performed a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, each species within an EEZ was classified based on the habitat preference [@REF] and catch trend [@FishForVisa]. Moreover, each EEZ was classified based on the continental [@REFUN] and biome [@Gabs] location. Finally, we also include the ToE of both bottom and sea surface temperature. We hypothesize that all of these factors would have significant influence in the time of emergence of transboundary stocks. Finally, we ran a series of ANOVAS to test continental and habitat preference ToE differences betwwn transboundary species.

All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *wesanderson* [@Packagewesanderson:2018vn], *rfishbase* [@PackagerfishbaseR:2019th], *R.matlab* [@PackageRmatlabRe:2018un], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated changes in the Stock Share Ratio of transboundary species shared by neighbouring EEZs by early (2020-2040) and mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). While most EEZs will experience an average change > 20%, a few EEZs are expected to have larger changes in SSR (Fig. 1A). For example, Guatemala' average SSR of *species* shared with Mexico is projected to go from 12% to 26% while Honduras SSR of *species* shared with the Nicaragua in the Pacific coast is projected to decline from 50% to 14%. The number of stocks per EEZ that will see a change in SSR by early $21^{st}$ century is relativelly low (*XX%*), however, some regions like West Africa and the Mediterranean will see changes in over 50 species. These regions, where EEZs are relativelly small and have multiple borders will see a complex dynamic of changes in SSR with no clear patterns (Fig. 2B). Towards the mid $21^st$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to early century, but are expected to increase in intensity (**Supplements Fig 2 (histogram)**).

@@

We estimated the time of emergence between neighbouring EEZs for 633 transbboundary stocks. Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least one neighbouring EEZ (e.g. Peruvean anchoveta emerged in 2010 between Chile and Peru). The cumulative sum of emerging stocks will have a steeper increase untill 2040 where the increment wil be slightly less intense. Similarly, for the number of EEZs, the increment curve seem to flattern by 2020. By 2081, all of the EEZs analyzed here will have at least one emerging stock while 2097 will be the year where all 633 stocks have emerged in at least one neighbouring EEZ (Fig. 2).

@@

On average, most of the world EEZ's will see a stock ToE by 2030 with a higher frfequency of number of emerging stocks between latitudes 0$^o$ and 50$^o$N (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50%, however, in some cases it represents 85% of total fishing revenue (Peru) or more, like the case of island territories such as Marshall Islands (92%) or Tokelau- New Zealand (100%). In few cases like Brazil's EEZ and Alaska's polar EEZ, no stock will emerge.

@@

The regional patterns of ToE suggest that most stocks will emerge by early 21st with many stocks in the Americas and Oceania already emerging. For the larger fisheries of the world in terms of revenue representig 75 percentile, ony the USA will see an average ToE before 2030 while fisheries in all other top fishing countries will be emerging within the following 10 years.

@@

We hypothesize that species environmental preferences (e.g. presence in sertain biomes and habitat), climate change effects on water temperature, human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the time of emergence of transboundary stocks (Table 1). Results from the multiple linear regression analysis show a significant relation between ToE of transbounuadry stocks and latitudinal gradient (*p* > 0.001), **biome2** (*p* = 0.006) and speicies habitat association (Table 1).

>>>>

The delineation of Economic Exclusive Zones (EEZs) by the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, established political boundaries across the distribution of many fish species, creating shared species between nations. Despite not implicit, UNCLOS incentivizes nations to cooperate in the management of transboundary fisheries and most management plans emplaced are based on historical distribution of catches. Climate change represents a challenge for the joint management of fish stocks. For instance, climate-driven changes in species distributions have led to conflict between nations due to changes in the proportion of captures, quota allocation, and fishery newcomers. Recent events suggests that even changes in species that contribute a relatively small proportion of the overall catch can lead to important international conflict when no legal framework is available. Thus, it is fundamental to prepare for climate change-driven impacts such as identifying the stocks that are going to change historical proportions within neighboring EEZs, the degree by which they will change, and the time when such change is going to happen. Only by adapting management plans and policy that anticipates to these effects will international fisheries treaties be climate-ready. Here, we build on previous work to identify the transboundary stocks that will see changes in their stock-share-proportion (e.g. the proportion of the stock that each EEZ has) around the world. Moreover, we determine the time of emergence of each fish stock, defined as the year in which the stock-share-proportion will change from its historical range. Our results suggest that by the mid of the 21st century EEZs will see changes in stock-share ratio between 1 and 10% with some exceeding 25%. Most of these changes will happen before 2050 urging policy to be re-formulated to account for the possible implications of these changes. By the end of the 21st century virtually all EEZs and transboundary species will have experienced changes in stock-share-proportion relative to historical distributions. The results of this research has the potential to improve climate-ready fisheries if policy is modify to respond to the consequences presented here.

@@

The world is moving towards sustainable development and ocean-based activities plays a mayor role in achieving international policies.

One of the main goals of the United Nations Sustainable Development Goal No.14 (SDG 14-Life below water) is to achieve sustainable development of fisheries [@SDG14].
<!-- (https://sustainabledevelopment.un.org/sdg14). -->

Climate change is expected to reduce the production from global fisheries in about XX% to xx% [@Cheung; @Barange] compromising our capacity to reach SDG 14.

Recent research show that the impacts of climate change on the productivity of marine wild fisheries can be offset with the proper reform of current fisheries management [@Gaines; @CheungClimateReadyFish]. While bringing fisheries that are poorly managed to a sustainable level will likely offset the negative impacts of climate change to fisheries production [@Gaines], it will not necessarily solve the spatial consequences of species moving from one management area to another.

Shifts in species distribution does not necessarily mean a reduction in production (e.g. reduction in stock biomass) but rather a spatial re-distribution of the stock's biomass [@RefLiam?].

Documented cases of stock shifts have resulted in international conflict and sustainable fishing rates by parties involved in the conflict.

This is of specific concern to a set of 633 fished species that cross international borders, also known as transponder species [@FishForVisa].

With the implementation of arbitrary spatial management boundaries, at least 633 species became transboundary representing an average catch (2005 - 2010) of 47.6 million tons and generating USD 65,483 million in fishing revenue [@FishorVisa].

Transboundary fish stocks refer to those species that cross international jurisdictions known as Economic Exclusive Zones (EEZs) of neighboring nations [@FAO]. These stocks are responsible for an average catch (2005 - 2010) of 47.6 million tons generating USD 65,483 million in fishing revenue [@FishorVisa].

Historically, environmentally-driven shits in shared stocks have cause problems and threaten stock's sustainability. **In early 1980's Canada and the US blablabla.** Many fish stocks managed by the European Union ICES regultaroy areas have been shifting since the 1970's [@EuropeShift]. Such shifts have already result in confrontations between nations as evidenced by the expansion of the Atlantic mackerel (*Scomber scombrus*) distribution reaching the EEZ of Iceland [@Spijkers:2017ij].  More recently, Chile and Peru have declare a cooperation strategy to mange the transboundary stock of Peruvean Anchovera [@Peru]. Furthermore, as we continue to enter the $21^{st}$ century, commercial marine species are expected to continue shifting (mostly poleward) potentilly compromising shared stocks management. By the end of the century, all EEZs of the world are proyected to have at least one new transboundary species with some of them having over 10 new stocks [@Pinsky]. Consequently, as species occupy new waters they leave historic areas and so, xx% of species are expected to exit EEZs by the end of che century [@Oremus]. This entering and exicting national jurisdicions will complicate fisheries management potentially creating confllict between neighbouring nations [@Pinksy @Oremus:2020id].

 So, preparing policy to species on the move is key to achieve sustainable development goals and proper governance of the world fisheries [@Pinksy].

Undrstanding the degree by which shared stocks are going to change and the time of emergence is key for creating climate-ready fisheries policy.

The concepts of transboundary stocks and EEZs were introduced in the early 1980's by the United Nations Law of the Sea (UCLOS) [@UnitedNations:1986tl]. Under UNCLOS each costal nation has the right (and responsabillity) to manage wild fish stocks up to 200 nautical mailes of the coast. Exceptions exist when territorial waters are in dispute (e.g. South China sea[@Ref]) or when menagement agreements exist between nations [@Ref]. In fact, UNCLOS incentives actions to cooperate on the management of shared stocks [Article 63, @UnitedNations:1986tl]. Climate change is shifting the distribution of marine species to the extreem that some spatial policies implemented to manage fisheries are becoming obsolete [@Pololanza; @Jessica].

@@

## Databases and species selection

The Food and Agricultural Organization (FAO) of the UN define transboundary stocks as those that freely swim across the EEZ of neighboring nations [@REFFAO]. Accordingly, we included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs [@Palacios-Abrantes:xxxx]. We consider each species as a single stock unit within an EEZ due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the *Sea Around Us* catch data to estimated the average catch and fishing revenue (USD 2019). We only used data from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we catalogue catch trends as increasing, constant and decreasing, following Palacios-Abrates *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Gabriel], and species were attributed a habitat preference [@FishBase]. For the present study we first projected the 633 species distributions under climate change and then estimated two main metrics; a stock's share ratio (SSR) and time of emergence (ToE) as described in the following sections.

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM). The DBEM algorithm includes species physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in an unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members [@Rodggers:2015jq].

## Changes in Stocks Share Ratio

In game theory, a threat point is known as the minimum payoff that a player is willing to receive in order to cooperate with other players. Thus, if a player's payoff is lower than the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S]. Here, we adopted the concept of threat point as the minimum proportion of the catch that each country will accept if they were to cooperate in the management of a transboundary stock [@Palacios-Abrantes:E&S] hereafter referred as stock share ratio (SSR). For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@Palacios-Abrantes:xxxx]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^st$ century as the average of 2021 to 2040 ($t_e$), and mid $21^st$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. We then estimate the percentage change in the SSR of each future time period relative to the historic time period. To address the natural spatial variability of the stock we only considered a SSR change when the future SSR was higher or lower than the average historic SSR $\pm$ 1 s.d. The analysis was replicated for each of the ten ensemble members capturing variation by averaging results among ensemble members (mean $\pm$ standard deviation).

## Time of Emergence Index

Knowing the time (e.g. year) in which SSR is going to diverge from a historical variability can be key for policy to prepare to such change. In here we call this event time of emergence and defined it as the moment in time when the future variation in SSR becomes larger than the historical trend variability [@Roggers:2015jq]. We employed the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

$$D_{ens} = acos(sin(lat_{eez})*sin(lat_{spp}) + cos(lat_{eez})*cos(lat_{spp})*cos(lon_{eez}-lon_{spp}))$$
where $lat_{eez}$ and $lat_{spp}$ are the latitudes of the EEZ and specie centroid, respectively and $lon_{eez}$ and $lon_{spp}$ are the respectively longitudes.

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*h*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, to determine the time of emergence we estimated a signal to noise ratio (SNR) and set an arbitrary threshold as follows [@Rodggers:2015jq]:

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated following the previous equation.

<!-- In cases where the species is distributed in more than 2 neighboring EEZs (e.g. *xxx*), the index is as follows: -->

@@

In other words, for two neighboring EEZs sharing a stock, if the centroid shifts towards EEZ A ($D_A$ decreases) but away from EEZ A B ($D_B$ increases), EEZ A may catch a bigger share over B and the risk of dispute increases (see *TI* equation). Consequently, if the centroid shifts are within historical variability (*SNR* < 1), risk of dispute between EEZ A and B is lower, finally, if the centroid shifts are beyond historical variability (*SNR* > 1), the risk of dispute is higher. Thus, risk is positively related to the frequency of centroid shifts beyond historical variability.

## Statistical analysis

We tested data for normallity (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the time of emergence of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we lookd at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the time of emergence of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated changes in the Stock Share Ratio of transboundary species shared by neighboring EEZs by early (2020-2040) and mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). While most EEZs will experience an average change > 20%, a few EEZs are expected to have larger changes in SSR (Fig. 1A). For example, Guatemala' average SSR of *species* shared with Mexico is projected to go from 12% to 26% while Honduras SSR of *species* shared with the Nicaragua in the Pacific coast is projected to decline from 50% to 14%. The number of stocks per EEZ that will see a change in SSR by early $21^{st}$ century is relatively low (*XX%*), however, some regions like West Africa and the Mediterranean will see changes in over 50 species. These regions, where EEZs are relatively small and have multiple borders will see a complex dynamic of changes in SSR with no clear patterns (Fig. 2B). Towards the mid $21^st$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to early century, but are expected to increase in intensity (**Supplements Fig 2 (histogram)**).

@@

We estimated the time of emergence between neighboring EEZs for 633 transboundary stocks. Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least one neighboring EEZ (e.g. Peruvean anchoveta emerged in 2010 between Chile and Peru). The cumulative sum of emerging stocks will have a steeper increase until 2040 where the increment will be slightly less intense. Similarly, for the number of EEZs, the increment curve seem to flatten by 2020. By 2081, all of the EEZs analyzed here will have at least one emerging stock while 2097 will be the year where all 633 stocks have emerged in at least one neighboring EEZ (Fig. 2).

@@

On average, most of the world EEZ's will see a stock ToE by 2030 with a higher frequency of number of emerging stocks between latitudes 0$^o$ and 50$^o$N (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50%, however, in some cases it represents 85% of total fishing revenue (Peru) or more, like the case of island territories such as Marshall Islands (92%) or Tokelau- New Zealand (100%). In few cases like Brazil's EEZ and Alaska's polar EEZ, no stock will emerge.

@@

The regional patterns of ToE suggest that most stocks will emerge by early 21st with many stocks in the Americas and Oceania already emerging. For the larger fisheries of the world in terms of revenue representing 75 percentile, only the USA will see an average ToE before 2030 while fisheries in all other top fishing countries will be emerging within the following 10 years.

@@

We hypothesize that species environmental preferences (e.g. presence in certain biomes and habitat preference), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary stocks (*Table Sup 3*). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary stocks and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (Table 1).
